Multiple Linear Regression on Sleep Data

### 2025-05-08 Kyubin's Edit/paper_KYUBIN.qmd

Next, we want to take care of multicollinearity after removing observation 264 from the data set. We checked the VIF values for each input covariate and determined which features can be removed. We also set 5 as a threshold for a large VIF value and removed any features that exceed this value. Furthermore, we used the scaled GVIF value, $\text{GVIF}^{1 / (2 \cdot \text{Df})}$, as it allows categorical variables to be numerically compared to continuous variables. From our VIF table, we noticed that `age` and `quality_of_sleep` have $\text{GVIF}^{1 / (2 \cdot \text{Df})} > 5$.

@@

full_model_test<-lm(sleep_duration ~., sleep_test)
summary(full_model_test)

@@

We removed variables `age` and `quality_of_sleep` and ran a MLR model. Our results showed that the results from reduced VIF model was in very similar to the full model. Adjusted R^2 was essentially unchanged. Therefore, we concluded that `age` and `quality_of_sleep` did not contribute much to the regression analysis because they were linear combinations of other independent variables.

```{r}
sleep_test_vif <- sleep_test %>%
  dplyr::select(!c(age, quality_of_sleep))

@@

model_stats_vif <- tibble::tibble(
  `F-stat.` = round(summary(full_model_vif)$fstatistic[1], 2),
  `Adjusted R²` = round(summary(full_model_vif)$adj.r.squared, 3),
  `Residual Sum of Squares` = round(sum(residuals(full_model_vif)^2), 2),
  `Total Sum of Squares` = round(sum((sleep_test_vif$sleep_duration - mean(sleep_test_vif$sleep_duration))^2), 2))

knitr::kable(model_stats_vif, caption = "Outlier + VIF Model")

@@

>>>>

Next, we wanted to take care of multicollinearity after removing observation 264 from the data set. We checked the VIF values for each input covariate and determined which features can be removed. We also set 5 as a threshold for a large VIF value and removed any features that exceed this value. Furthermore, we used the scaled GVIF value, $\text{GVIF}^{1 / (2 \cdot \text{Df})}$, as it allows categorical variables to be numerically compared to continuous variables. From our VIF table, we noticed that `age` and `quality_of_sleep` have $\text{GVIF}^{1 / (2 \cdot \text{Df})} > 5$.

@@

full_model_test<-lm(sleep_duration ~., sleep_test)

@@

We removed variables `age` and `quality_of_sleep` and ran a MLR model. Our results showed that the reduced VIF model was very similar to the full model. Adjusted R^2 was essentially unchanged. Therefore, we concluded that `age` and `quality_of_sleep` did not contribute much to the regression analysis because they were linear combinations of other independent variables.

```{r include=FALSE}

sleep_test_vif <- sleep_test %>%
  dplyr::select(!c(age, quality_of_sleep))

@@

@@

```{r}
model_stats_vif <- tibble::tibble(
  `F-stat.` = round(summary(full_model_vif)$fstatistic[1], 2),
  `Adjusted R²` = round(summary(full_model_vif)$adj.r.squared, 3),
  `Residual Sum of Squares` = round(sum(residuals(full_model_vif)^2), 2),
  `Total Sum of Squares` = round(sum((sleep_test_vif$sleep_duration - mean(sleep_test_vif$sleep_duration))^2), 2))

knitr::kable(model_stats_vif, caption = "Outlier + VIF Model")
```

\noindent\textbf{\Large Variable Selection: Backward, Forward, and Stepwise}

In order to avoid overfitting issues, we utilized a greedy algorithmn process to select variables that provided the lowest AIC value. We ran 3 model selection process as discussed in class and recorded the variables selected for each method. We found that all 3 methods took all of our variables in our data set. We knew that AIC penalizes the fitted model for each additional variables taken. Despite this, we saw that all variables were significant. Note that the variable selection process was done after our high VIF variables were removed.

```{r include=FALSE}
# Load required packages
library(knitr)

# 1. Model Selection using full_model_vif
model_backward <- step(full_model_vif, direction = "backward")

# Define full model formula and null model for forward/stepwise
full_scope <- formula(full_model_vif)
null_model <- lm(sleep_duration ~ 1, data = sleep)

model_forward <- step(null_model, scope = full_scope, direction = "forward")
model_stepwise <- step(null_model, scope = full_scope, direction = "both")

# 2. Extract variable names selected by each model
vars_backward <- attr(terms(model_backward), "term.labels")
vars_forward <- attr(terms(model_forward), "term.labels")
vars_stepwise <- attr(terms(model_stepwise), "term.labels")

# 3. Combine all variables that appeared in any model
all_vars <- sort(unique(c(vars_backward, vars_forward, vars_stepwise)))

# 4. Create a comparison table
selection_table <- data.frame(
  Variable = all_vars,
  Backward = ifelse(all_vars %in% vars_backward, "Yes", "No"),
  Forward  = ifelse(all_vars %in% vars_forward, "Yes", "No"),
  Stepwise = ifelse(all_vars %in% vars_stepwise, "Yes", "No")
)

```{r}

kable(selection_table, caption = "Variable Selection")

\noindent\textbf{\Large Suggested Regression: LASSO}

For our suggested regression, we used LASSO regression. We hoped to extract the most import input features to achieve a simple yet powerful model. To find the optimal $\lambda$ or the regularization parameter, we used a cross-validation process by splitting our data set: train = 80%. We saw that `gender`, `occupation`, `bmi_category`, and `sleep_disorder` converged to zero.

```{r include=FALSE}

library(glmnet)
library(tidyverse)
library(tidymodels)
library(tidyr)
library(car)
library(caret)

## Split Training/Test Data
sleep_sp <- sleep_test_vif %>% initial_split(prop = 0.8, strata=sleep_duration)
rtrain <- sleep_sp %>% training()
rtest <- sleep_sp %>% testing()

model0 <- lm(sleep_duration ~ ., data=rtrain)
summary(model0)

## Regularization
xrtrain <- rtrain %>% dplyr::select(-sleep_duration) %>% as.matrix()
yrtrain <- rtrain$sleep_duration

xrtest  <- rtest %>% dplyr::select(-sleep_duration) %>% as.matrix()
yrtest  <- rtest$sleep_duration

modelL <- glmnet(xrtrain, yrtrain)
plot(modelL,label = TRUE)
## Df=# of nonzero coefficients, %Dev = % of Deviance explained (like SSR)
print(modelL)
## Coefficient for a certain lambda
coef.glmnet(modelL, s= 0.1)

## Cross-Validation
## nfolds = # of groups ex) 20
## Regression with 19 groups and test with 1 group
cvmodelL <- cv.glmnet(xrtrain, yrtrain, type.measure = "mse", nfolds = 20)
coef.glmnet(modelL, s= cvmodelL$lambda.min)
plot(cvmodelL)

coef_min_lambda <- coef.glmnet(modelL, s = cvmodelL$lambda.min)

# Convert to a tidy data frame and filter out zero coefficients
nonzero_coefs <- as.matrix(coef_min_lambda)
nonzero_coefs_df <- data.frame(
  Variable = rownames(nonzero_coefs),
  Coefficient = nonzero_coefs[, 1]
)

```{r}
kable(nonzero_coefs_df, digits = 4, caption = "Non-Zero Coefficients from LASSO (Selected by Cross-Validation)")
```
